---
title: "Statistical Inference Part B"
author: "Suhas P K"
date: "2023-06-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringi)
```
## Project 2 : Basic inferential data analysis

Now in the second project, we're going to analyze the ToothGrowth data in the R datasets package.

1. Load the ToothGrowth data and perform some basic exploratory data analyses

2. Provide a basic summary of the data.

3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

4. State your conclusions and the assumptions needed for your conclusions.

### About Tooth growth dataset.
The **Tooth growth data set** is one of the standard learning data set included in R. The tooth growth data set is the length of the **ondoblasts (teeth)** in each of 10 guinea pigs at three Vitamin-C dosage levels (0.5,1,and 2 mg) with two delivery methods (orange juice or ascorbic acid).

The file contains 60 observations of 3 variables,
- len : Tooth length
- supp : Supplement type (VC or OJ)
- dose : Dose in milligrams

### Data analysis of the **Tooth Growth data**
```{r libraries,echo=FALSE}
if (!require(stats)){
    install.packages("stats")
    library(stats)
}
if (!require(ggplot2)){
    install.packages("ggplot2")
    library(ggplot2)
}
if (!require(ggdark)){
    install.packages("ggdark")
    library(ggdark)
}

```
```{r data table}
head(ToothGrowth, 10)
```

Load the data set.
```{r tooth_growth_data,results='asis'}
data(ToothGrowth)
```
```{r tooth length vs dose, echo=FALSE}
qplot(dose, len,
      data = ToothGrowth,
      color = supp,
      geom = "point")+
    geom_smooth(method = "lm")+
    labs(title = "Tooth Growth")+
    labs(x = "Dose of supplements",
         y = "Length of teeth")+
    dark_theme_light()
```
**Analysis from the plot**
1. The length of teeth goes up as the does of supplements increase, which indicates that the supplements may help teeth growth.
2. At the same dose, OJ seems to incur a higher of teeth growth than VC.
3. The slope of OJ is not as steep as the slope of VC, meaning an increase in VC may make a larger increase in teeth length than in OJ.

```{r candle plot}
qplot(supp,len,data=ToothGrowth,
      facets=~dose, main="Tooth growth of guinea pigs by supplement type and dosage (mg)",
      xlab="Supplement type", 
      ylab="Tooth length") + 
    geom_boxplot(aes(fill = supp)) + dark_theme_grey()
```
There is a positive effect of the dosage, as the dosage increases the tooth growth increase. In the specific case of the VC, the tooth growth has a linear relationship with dosage. The higher dosage (2.0 mg) has less improvement in tooth growth with the OJ supplement. However, the OJ supplement generally induces tooth more growth than VC except at higher dosage (2.0 mg).

**Hypothesis Testing**

Assumptions
- The variables must be independent and identically distributes (iid).
- Variances of tooth growth are different when using different supplement and dosage.
- Tooth growth follows a normal distribution.

**Hypothesis for the supplement OJ vs VC**

Let our null hypothesis to be there is no difference in tooth growth when using the supplement OJ and VC.

Let our alternate hypothesis to be there more tooth growth when using supplement OJ than VC.
```{r supplement name1}
OJ <- ToothGrowth$len[ToothGrowth$supp == 'OJ']
VC <- ToothGrowth$len[ToothGrowth$supp == 'VC']
```
Let us perform a t-test following the indications of the work to be evaluated. \
One-tailed independent t-test with unequal variance.
```{r}
t.test(
    OJ,VC,
    alternative = "greater",
    paired = FALSE,
    var.equal = FALSE,
    conf.level = 0.95
)
```

As the p-value (0.03032) is lower than 0.05 (the default value for the tolerance of the error alpha), then, we **reject the null hypothesis**. This can be interpreted as there is approximately 3% of chance to obtain an extreme value for the difference in mean of the tooth growth.

Based on this low p-value, it can be concluded that it is very likely that supplement OJ, the greater the effect on tooth growth than supplement VC.

**Hypothesis for dosage.**

The null hypothesis is that there is no difference in tooth growth between dosage. Our alternate hypothesis is that there are most tooth growth when the dosage increases.

Extract the tooth growth by dosage.
```{r doseage}
dose_half <- ToothGrowth$len[
    ToothGrowth$dose == 0.5
]
dose_one <- ToothGrowth$len[
    ToothGrowth$dose == 1
]
dose_two <- ToothGrowth$len[
    ToothGrowth$dose == 2
]
```

One-tailed independent t-test with unequal variance.
```{r t-test for dose_half_one}
t.test(dose_half,
       dose_one,
       alternative = "less",
       paired = FALSE,
       var.equal = FALSE,
       conf.level = 0.95)
```

As the p-value (6.342e-08) is lower than 0.05 (the default value for the tolerance of the error alpha), then, we **reject the null hypothesis**. That can be interpreted as there is almost null chances of obtain an extreme value for the difference in mean of those dosages (dose_half < dose_one) on the tooth growth.

```{r t-test for does_one_two}
t.test(
    dose_one,dose_two,
    alternative = "less",
    paired = FALSE,
    var.equal = FALSE,
    conf.level = 0.95
)
```

The conclusion is similar than the previous, the p-value is 9.532e-06, close to 0. Then we reject the null hypothesis. That can be interpreted as there is almost null chances of obtain an extreme value for the difference in mean of those dosages (dose_one < dose_two) on the tooth growth. The value is extreme (that’s what we reject the null hypothesis)

Based on these low p-values, we can conclude that it is very likely that dosage has effect, and a higher dosage higher tooth growth.

**Hypothesis for the supplement OJ vs VC at dosage 2.0 mg**

```{r supplement name2,echo=FALSE}
OJ2 <- ToothGrowth$len[
    ToothGrowth$supp == 'OJ' &
        ToothGrowth$dose == 2
]
VC2 <- ToothGrowth$len[
    ToothGrowth$supp =='VC' &
        ToothGrowth$dose ==2
]
```

Two-tailed independent t-test with unequal variance.

```{r t-test for supp}
t.test(OJ2, VC2, 
       alternative = "two.sided", 
       paired = FALSE, 
       var.equal = FALSE, 
       conf.level = 0.95)
```

The p-value (0.9639) confirm what we suspect, that we can’t reject the null hypothesis (p-value is higher than 0.05 (the default value for the tolerance of the error alpha). Then, there is insufficient evidence to show that there is a difference in tooth growth when using supplement OJ and VC at dosage 2.0 mg.

**CONCLUSION**

The conclusion is when the dose is 0.5 or 1.0 there is a difference between the teeth growth after taking OJ and VC, while when the dose is 2.0, there is no difference between the teeth growth after taking OJ and VC. The assumption needed is we first assumed the whole population is normally distributed, then we assumed the population is normally distributed under each dose.

